Geometry of Ranked Nearest Neighbour Interchange Space of Phylogenetic Trees

### 2019-12-12 rnni_convexity.tex
fixed proof of FindPath correctness

\begin{algorithm}[H]
\caption{$\findpath$($T,R$)}
\label{alg:find_path}
\begin{algorithmic}[1]
\STATE $T' := T$, $p := [T']$
\FOR {$k = 1, \dots, n-2$}
  \STATE Let $C_k$ be the cluster induced by the node with rank $k$ in $R$ \label{alg:find_path:line:cluster}
  \WHILE {$\rank_{T'}(\mrca_{T'}(C_k))>k$}
    \STATE Update $T'$: Decrease $\rank_{T'}(\mrca_{T'}(C_k))$ by an $\rnni$ move \label{alg:findpath:line:move_set_down}
    \STATE $p = p+T'$
  \ENDWHILE
\ENDFOR
\RETURN $p$
\end{algorithmic}
\end{algorithm}

The following lemma shows that $\findpath$ is a correct deterministic algorithm.

\begin{lemma}
Let $T = [C_1, \ldots, C_{k-1}, B_{k}, \ldots, B_{n-2}]$ be a tree and $C_k$ a cluster such that $\mrca_T(C_k) > k$.
Then there exists exactly one $\rnni$ move on $T$ that results in a tree $T' = [C_1, \ldots, C_{k-1}, \ldots]$ with $\mrca_{T'}(C_k) < \mrca_{T}(C_k)$.
\label{lemma:mrca_move}
\end{lemma}

\todo{We never use $R$ in this proof, so $B_k$'s are irrelevant in Lemma.}
\begin{proof}
Let $T$ and $R$ be trees as defined in the lemma.
Notice that the only $\rnni$ move on $T$ that could decrease the rank of $\mrca_T(C_k)$ are $\nni$ moves on the edge between $\mrca_T(C_k)$ and its child (if this edge has length one), or a rank move between $\mrca_T(C_k)$ and the node with rank one less.
Therefore, we will now distinguish these two cases:
Either there is an edge $(v,w)$ between nodes $v = \mrca_T(C_k)$ and $w$ with rank $rank_T(w) = rank_T(v) - 1$, or there is no such edge.

In the latter case the only $\rnni$ move that could decrease the rank of $v$ is a rank move swapping the ranks of $v$ and $w$.

Let us now assume that there is an edge $(v,w)$ in $T$.
Let $T_v$ be the subtree that is rooted in the child of $v$ that is not $w$ and let $T_{w_1}$ and $T_{w_2}$ be the subtrees rooted in the children of $w$.
We are now going to explain why only one of these two $\nni$ moves changes the rank of $\mrca_T(C_k)$.
It is important to notice that $C_k$ either is the disjoint union of two of the clusters $C_1, \ldots, C_{k-1}$ or the union of one of those clusters with a single taxon that is not included in any of those clusters.
In general we can say that $C_k = S_1 \dot\cup S_2$ is the union of two subsets of taxon set of $T$ such each of these subsets is contained in one of the subtrees $T_v, T_{w_1}$, or $T_{w_2}$.
In fact, $S_1$ and $S_2$ are contained in two different of these subtrees.
Now it is easy to see that just one of the $\nni$ moves on $(v,w)$ (illustrated in Figure~\ref{fig:mrca_move}) decreases the rank of $\mrca_T(C_k)$.

\begin{figure}[H]
\centering
\includegraphics[width=0.5\textwidth]{mrca_move}
\vspace{12pt}
\caption{Two $\nni$ moves that are possible on the edge $(v,w)$ of the tree $T'$ with subtrees $T_v, T_{w_1}, T_{w_2}$ as described in the proof of Lemma~\ref{lemma:mrca_move}.}
\label{fig:mrca_move}
\end{figure}
\end{proof}

@@

The algorithm $\csort$ (Algorithm~\ref{alg:csort}) is a natural modification of the classical Bubble Sort algorithm \autocite{Knuth1997-pi}.
A path $p$ from $T$ to $R= [x_1, x_2, \ldots, x_n]$ is computed iteratively so that after $k$ steps the last $k$ taxa of $T$ and $R$ coincide.
Specifically, in step $k$ ($k = n, \ldots, 3$) the parent of taxon $x_{k}$ is moved up by $\nni$ moves to the position it has in $R$.
Notice that there might be more than just one such move per step, which means that $p$ can be extended by more than one tree in each step.

>>>>

\begin{algorithm}[H]
\caption{$\findpath$($T,R$)}
\begin{algorithmic}[1]
\STATE $T' := T$, $p := [T']$, $[C_1, \ldots, C_{n-2}] := R$
\FOR {$k = 1, \dots, n-2$}
  \WHILE {$\rank_{T'}(\mrca_{T'}(C_k))>k$}
    \STATE Update $T'$: Decrease $\rank_{T'}(\mrca_{T'}(C_k))$ by an $\rnni$ move
    \label{alg:findpath:line:move_set_down}
    \STATE $p = p+T'$
  \ENDWHILE
\ENDFOR
\RETURN $p$
\end{algorithmic}
\end{algorithm}

\begin{proposition}
$\findpath$ is a correct deterministic algorithm.
\end{proposition}

\proof
It is suffice to show that the update operation (line~\ref{alg:findpath:line:move_set_down}) is well-defined, that is the $\rnni$ move that decreases the rank of $\mrca_{T'}(C_k)$ is unique.

Case $k = 1$.
In this case $C_k$ is a cherry $(x, y)$.
The node $v = \mrca_{T'}(x, y)$ has $\rank_{T'}(v) = r > 1$.
Consider the node $w$ preceding $v$ in $T'$, that is, $\rank_{T'}(w) = r - 1$.
If a rank move that swaps $v$ and $w$ is possible then this is the only move on $T'$ that can decrease the rank of $\mrca(x, y)$.
If the rank move is impossible then there is an edge in $T'$ connecting $v$ and $w$.
The only way to decrease the rank of $\mrca(x, y)$ then is to perform an $\nni$ move on that edge.
It is not hard to see that out of two possible $\nni$ moves only one decreases the rank of $\mrca$.

Case $k > 1$.
\begin{enumerate}
\item $C_k = C_i \cup C_j$ for $i, j < k$.
Suppress $C_i$ and $C_j$ in both $T'$ and $R$ to new taxa $c_i$ and $c_j$ respectively and proceed as in Case $k = 1$.
\item $C_k = C_i \cup {x}$, where $x$ is a taxon not present in $C_1, \ldots, C_k$.
Suppress $C_i$ in both $T'$ and $R$ to a new taxon $c_i$ and proceed as in Case $k = 1$.
\item $C_k = \{x, y\}$ where both $x$ and $y$ are not in $C_1, \ldots, C_k$.
Proceed as in Case $k = 1$.
\end{enumerate}
\endproof

@@

The algorithm $\csort$ (Algorithm~\ref{alg:csort}) is a modification of the classical Bubble Sort algorithm \autocite{Knuth1997-pi}.
A path $p$ from $T$ to $R= [x_1, \ldots, x_n]$ is computed iteratively so that after $k$ steps the last $k$ taxa of $T$ and $R$ coincide.
Specifically, in step $k$ ($k = n, \ldots, 3$) the parent of taxon $x_{k}$ is moved up by $\nni$ moves to the position it has in $R$.
Notice that there might be more than just one such move per step, which means that $p$ can be extended by more than one tree in each step.
